---
title: "09 - Rolling Functions with Missing Values"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{09 - Rolling Functions with Missing Values}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
editor:
  markdown:
    wrap: sentence
bibliography: ["references.bib"]
---

## Notes

* These functions ignore `NA` values for now. Adjustments for handling `NA` values are covered in a separate vignette.
* R already provides efficient versions of the functions covered here. This is just to illustrate how to use C++ code.

## Cumulative sum

The following function expands the previous `cumsum_cpp()` function to handle missing values.

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

```cpp
[[cpp4r::register]]
doubles cumsum2_cpp(doubles x, bool na_rm = false) {
  int n = x.size();

  writable::doubles out(n);
  out[0] = x[0];

  if (na_rm == true) {
    for (int i = 1; i < n; ++i) {
      double y1 = out[i - 1], y2 = x[i];
      if (ISNAN(y2)) {
        out[i] = y1 + 0.0;
      } else {
        if (ISNAN(y1)) {
          out[i] = 0.0 + y2;
        } else {
          out[i] = y1 + y2;
        }
      }
    }
  } else {
    for (int i = 1; i < n; ++i) {
      double y1 = out[i - 1], y2 = x[i];
      if (ISNAN(y2)) {
        out[i] = NA_REAL;
      } else {
        if (ISNAN(y1)) {
          out[i] = NA_REAL;
        } else {
          out[i] = y1 + y2;
        }
      }
    }
  }

  return out;
}
```

To test the functions, you can run the following benchmark code in the R console:

```{r cumsum-bench}
set.seed(123) # for reproducibility
x <- rpois(1e6, lambda = 2) # 1,000,000 elements

cumsum(c(1, NA, 2, 4))
cumsum2_cpp(c(1, NA, 2, 4))
cumsum2_cpp(c(1, NA, 2, 4), na_rm = TRUE)

mark(
  cumsum(x),
  cumsum2_cpp(x)
)
```

## Cumulative product

The following function expands the previous `cumprod_cpp()` function to handle missing values.

```cpp
[[cpp4r::register]]
doubles cumprod2_cpp(doubles x, bool na_rm = false) {
  int n = x.size();

  writable::doubles out(n);
  out[0] = x[0];

  if (na_rm == true) {
    for (int i = 1; i < n; ++i) {
      double y1 = out[i - 1], y2 = x[i];
      if (ISNAN(y2)) {
        out[i] = y1 * 1.0;
      } else {
        if (ISNAN(y1)) {
          out[i] = 1.0 * y2;
        } else {
          out[i] = y1 * y2;
        }
      }
    }
  } else {
    for (int i = 1; i < n; ++i) {
      double y1 = out[i - 1], y2 = x[i];
      if (ISNAN(y2)) {
        out[i] = NA_REAL;
      } else {
        if (ISNAN(y1)) {
          out[i] = NA_REAL;
        } else {
          out[i] = y1 * y2;
        }
      }
    }
  }

  return out;
}
```

To test the functions, you can run the following benchmark code in the R console:

```{r functions-cumprod-benchmark}
cumprod(c(1, NA, 2, 4))
cumprod2_cpp(c(1, NA, 2, 4))
cumprod2_cpp(c(1, NA, 2, 4), na_rm = TRUE)

mark(
  cumprod(x),
  cumprod2_cpp(x)
)
```

## Cumulative minimum

The next function calculates the cumulative minimum of the elements of a vector.

```cpp
[[cpp4r::register]] doubles
cummin_cpp(doubles x, bool na_rm = false) {
  int n = x.size();

  writable::doubles out(n);
  out[0] = x[0];

  if (na_rm == true) {
    for (int i = 1; i < n; ++i) {
      double y1 = x[i - 1], y2 = x[i];
      if (ISNAN(y1)) {
        out[i] = y2;
      } else {
        out[i] = std::min(y1, y2);
      }
    }
  } else {
    for (int i = 1; i < n; ++i) {
      double y1 = out[i - 1], y2 = x[i];
      if (ISNAN(y2)) {
        out[i] = NA_REAL;
      } else {
        if (ISNAN(y1)) {
          out[i] = NA_REAL;
        } else {
          out[i] = std::min(y1, y2);
        }
      }
    }
  }

  return out;
}
```

To test the function, you can run the following benchmark code in the R console:

```{r cummin-correctness}
cummin(c(1, NA, 2, 4))
cummin_cpp(c(1, NA, 2, 4))
cummin_cpp(c(1, NA, 2, 4), na_rm = TRUE)

mark(
  cummin(x),
  cummin_cpp(x)
)
```

## Cumulative maximum

The next function calculates the cumulative maximum of the elements of a vector.

```cpp
[[cpp4r::register]]
doubles cummax_cpp(doubles x, bool na_rm = false) {
  int n = x.size();

  writable::doubles out(n);
  out[0] = x[0];

  if (na_rm == true) {
    for (int i = 1; i < n; ++i) {
      double y1 = out[i - 1], y2 = x[i];
      if (ISNAN(y1)) {
        out[i] = y2;
      } else {
        out[i] = std::max(y1, y2);
      }
    }
  } else {
    for (int i = 1; i < n; ++i) {
      double y1 = out[i - 1], y2 = x[i];
      if (ISNAN(y2)) {
        out[i] = NA_REAL;
      } else {
        if (ISNAN(y1)) {
          out[i] = NA_REAL;
        } else {
          out[i] = std::max(y1, y2);
        }
      }
    }
  }

  return out;
}
```

To test the functions, you can run the following benchmark code in the R console:

```{r cummax-correctness}
cummax(c(1, NA, 2, 4))
cummax_cpp(c(1, NA, 2, 4))
cummax_cpp(c(1, NA, 2, 4), na_rm = TRUE)

mark(
  cummax(x),
  cummax_cpp(x)
)
```
